Sequence Specific Modeling of Cell-Free Protein Synthesis

### 2017-01-05 Paper_v1.tex
Nicks Edits

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The abstract environment will automatically gobble the contents
%% if an abstract is not used by the target journal.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
In this study, we used sequence specific constraints based modeling to evaluate the performance of synthetic circuits in an \emph{E.~coli} cell-free protein synthesis system.
A core \emph{E.~coli} metabolic model, consisting of glycolysis, pentose phosphate pathway, amino acid biosynthesis and degradation and energy metabolism, was then augmented with sequence specific descriptions of genetic circuits which included mechanistic models of promoter function, transcription and translation.
Thus, unlike other synthetic biology modeling efforts, sequence specific constraints based modeling explicitly couples the transcription and translation of circuit components with the availability of metabolic resources.
Model parameters for transcription and translation were taken from literature, allowing a first principles prediction of circuit performance.
We tested this approach by first simulating T7 induced CAT production and $\sigma_{70}$-induced deGFP expression; we then expanded these studies for a range of different proteins
First principles predictions of circuit performance were consistent with measurements for a variety of cases.
Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance in terms of productivity and yield.
A sufficient energy supply with oxidative phosphorylation is instrumental for high carbon yields, in addition the translation rate is the next step to optimize for higher productivity and yield.
Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.
\end{abstract}

@@

In this study, we used sequence specific constraints based modeling to evaluate the performance of synthetic circuits in an \emph{E.~coli} cell-free protein synthesis (CFPS) system.
A core \emph{E.~coli} cell free metabolic model, consisting of 132 metabolites and 281 reactions, was developed from literature \cite{Feist:2007aa}.
This model, which described glycolysis, pentose phosphate pathway, amino acid biosynthesis and degradation and energy metabolism, was then augmented with
sequence specific descriptions of genetic circuits which included mechanistic models of promoter function, transcription and translation.
Thus, sequence specific constraints based modeling explicitly coupled the transcription and translation of circuit components with the availability of metabolic resources.
We tested this approach by first simulating the production of CAT and deGFP under two different promoters. and then expanded this for $\sigma_{70}$-induced deGFP expression at different plasmid concentrations.
First principles predictions of circuit performance were consistent with measurements.
We then investigated the productivity and carbon yield for a range of 8 different proteins for a variety of cases.
From this, higher carbon number proteins typically had lower productivity rates and carbon yields than that of the lower carbon number proteins.
Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance, showing oxidative phosphorylation is instrumental for maintaining a high carbon yield and the translation rate for productivity.
Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell free synthetic circuits.

@@

\subsection{Model Derivation}
The goal of this work was first to construct a modeling framework to describe cell-free protein synthesis systems and to examine its performance in productivity and yield for a protein of interest.
One mathematical framework that has found wide use in modeling metabolism are constraint based models such as flux balance analysis (FBA).
FBA can predict how cells utilize nutrients to produce products by using the biochemical stoichiometry and thermodynamical feasibility under pseudo steady-state conditions.
Traditionally, FBA is used to model \textit{in vivo} processes, however cell-free systems do not have growth associated reactions or transport through the cell membrane.
Thus, to model cell-free metabolism we constructed a cell-free stoichiometric network by removing growth associated reactions from the MG1655 reconstruction \cite{Feist:2007aa}, and incorporating amino acid synthesis and transcription/translation associated reactions \cite{Allen:2003aa} for a protein of interest to be expressed.
The network consisted of 281 reactions and 132 species.
We developed a cell-free sequence specific flux balance analysis (ssFBA) with a detailed promoter model \cite{Moon:2012aa} to examine the performance of CFPS.
We first validated the ssFBA approach by comparing simulated and measured concentrations of two proteins from two different cell-free \textit{E. coli} extract systems.
The first protein, chloramphenicol acetyltransferase (CAT), was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress} for a duration of 1 hour under glucose consumption (Fig \ref{fig:CAT_GFP_prod}A).
The second protein, dual emission green fluorescent protein (deGFP), was produced under a P70 promoter in TXTL 2.0 \textit{E. coli} extract for a duration of 8 hours under maltose consumption (Fig \ref{fig:CAT_GFP_prod}B).
The ssFBA simulations predicted the production of both proteins for the duration of both CFPS batch reactions.
Uncertainty in experimental factors such as RNA polymerase, ribosome concentrations, elongation rates or the upper bounds for oxygen and glucose consumption rates did not alter the qualitative performance of the model.
Thus, the metabolic network and molecular description of trancription and translation were consistent with experimental measurements.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/CAT_GFP_prod.pdf}
\caption{Sequence specific flux balance analysis of protein production vesus time. A. CAT production under a T7 promoter in CFPS \textit{E. coli} extract for 1 hr under glucose consumption. B. deGFP production under a P70 promoter in TXTL 2.0 \textit{E. coli} extract for 8 hr under glucose consumption. 95\% CI (blue region) over the ensemble of 100 sets.}
\label{fig:CAT_GFP_prod}
\end{figure}

@@

\subsection{CFPS Productivity}
Our next goal was to examine the performance of CFPS productivity for eight different proteins under three different cases. Each of the proteins were produced under a P70 promoter, except for CAT which was produced under a T7 promoter. In all our cases, CFPS is supplied with glucose. The first case, CFPS is supplied with amino acids and the system can synthesize amino acids (control). In the second case, CFPS is supplied with amino acids, however the system cannot synthesize amino acids (AA uptake w/o synthesis). We turned off these synthesis reactions since during the cell-free extract preparation the cells are often supplied with amino acids, thus the enzymes responsible for amino acid synthesis would not be present. In the third case, CFPS is not supplied with amino acids, but the system can synthesize them (AA synthesis w/o uptakte).
Eight different proteins,ranging in size, were selected to evaluate CFPS performance: bone morphogenetic protein 10 (BMP10), chloramphenicol acetyltransferase (CAT), caspase 9 (CASP9), dual emission green fluorescent protein (deGFP), prothrombin (FII), coagulation factor X (FX), fibroblast growth factor 21 (FGF21), and single chain variable fragment R4 (scFvR4).
We used ssFBA to estimate the productivity for each of these proteins for each case (Fig \ref{fig:Prod_POI}A).
The first two cases had very similar performance, with the second case (without amino acid synthesis) having a slightly higher productivity for most proteins. 
%The second case (without amino acid synthesis) showed the highest productivity for each of the proteins, however the control case had very similar performance.
This shows the system had sufficient substrates to power the system and synthesize each protein.
The third case had the lowest productivity for each protein, thus the addition of amino acids to CFPS extract is important for maintaining a relatively high productivity.
The drop in productivity in the third case is due to the fact that glucose is the only substrate supplied and must be used to provide the necessary energy requirements for transcription and translation as well as synthesize each amino acids required for the production of the protein of interest.  
The qualitative trend of productivity for the three different cases was the same within each case, however some proteins had higher productivity than others.
For instance, in the second case FGF21 (fibroblast growth factor 21) had a productivity of 17 ($\mu$M/h) whereas FII (prothrombin) had a productivity of 3 ($\mu$M/h).
To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig \ref{fig:Prod_POI}B).
The proteins with the highest productivity had the lowest carbon number whereas proteins with low productivity had higher carbon numbers.
This inverse qualitative trend was due to the fact that larger proteins require more amino acids and substrates to assemble them resulting in lower productivity.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/Prod_POI.pdf}
\caption{CFPS productivity performace for eight proteins for the control (black), AA uptake w/o synthesis (off white), and AA synthesis w/o uptake (grey). A. Productivity normalized to the specific glucose uptake rate (Error bars represent the 95\% CI of the ensemble). B. Mean productivity versus the carbon number for each corresponding protein.}
\label{fig:Prod_POI}
\end{figure}

\subsection{CFPS Carbon Yield}
Following the same outline as in examining the productivity, we calculated the carbon yield for each protein (Fig \ref{fig:Yield_POI}A).
The same trends followed, where the case without amino acid synthesis showed the highest carbon yield and the control case had comparable performance.
The third case (with no amino acid uptake) had the lowest yields; this is most likely because glucose is utilized to synthesize the necessary amino acids for each protein as well as power the system. 
To determine the carbon contribution from each substrate (glucose and amino acids) we examined the carbon flux going toward the production of deGFP for all three cases (Table \ref{tbl:yield_breakdown}).
For the first case, the system is relied on a mixture of glucose and some amino acids for the productin of deGFP with a carbon yield of 20.5\%.
Once amino acid synthesis is removed from the network (2nd case), each amino acid is utilized and the carbon yield increased to 22.2\%.
Only the necessary amount of amino acids were used for the production of deGFP, thus it may be hypothesized that all the glucose was used to power CFPS and does not contribute to the carbon yield.
If that is the case, the carbon yield without glucose contribution is 100\%, whereas for the first case its 92.5\%.
Finally for the third case (without amino acid supplemented), the carbon yield was reduced to 14.8\% and the system used about double the amount of glucose as the first two cases.
In this case glucose is being used to synthesize amino acids and provide the necessary energy requirements to power transcription and translation; this trend was seen across all proteins except for CAT.
Interestingly, CAT carbon yield showed similar performance for all three cases, however it still had a lower yield than the first two cases.
Each protein has different energy requirements for transcription and translation depending on its sequence.
Thus, for the case of CAT, energy requirements for transcription or translation were low enough that it's carbon yield was not hindered in the third case as significantly as for the other proteins.

\begin{table}[t!]
\centering
    \caption{Carbon contribution from glucose and each amino acid toward deGFP production for three cases: control, amino acid uptake without synthesis, and amino acid synthesis without uptake.}
    \renewcommand{\arraystretch}{0.9}
    \begin{tabular}{lccc} \toprule
        \parbox{3.2 cm}{Carbon \phantom{abcdefghi} Produced (mM)} & Control & \parbox{2.5 cm}{AA uptake \phantom{ai} w/o synthesis} & \parbox{2.5 cm}{AA synthesis w/o uptake} \\ \midrule
        deGFP & 9.8 & 9.6 & 9.9  \\ \toprule
        \parbox{3.2 cm}{Carbon \phantom{abcdefghi} Consumed (mM)} & \\ \midrule
        GLC & 37.3 & 33.7 & 66.9 \\
        ALA & 0.0 & 0.2 & -  \\
        ARG & 0.3 & 0.3 & - \\
        ASN & 0.5 & 0.4 & -  \\
        ASP & 0.4 & 0.6 & -  \\
        CYS & 0.2 & 0.1 & -  \\
        GLU & 2.1 & 0.6 & -  \\
        GLN & 0.4 & 0.3 & - \\
        GLY & 0.3 & 0.3 & -  \\
        HIS & 0.5 & 0.5 & - \\
        ILE & 0.0 & 0.6 & -  \\
        LEU & 1.0 & 1.0 & - \\
        LYS & 1.0 & 0.9 & - \\
		MET & 0.0 & 0.2 & -  \\
        PHE & 1.0 & 0.9 & - \\
        PRO & 0.4 & 0.4 & -  \\
        SER & 0.0 & 0.2 & - \\
        THR & 1.0 & 0.5 & -  \\
        TRP & 0.1 & 0.1 & -  \\
        TYR & 0.8 & 0.8 & - \\
        VAL & 0.6 & 0.7 & - \\ \hline
        Sum & 47.9 & 43.3 & 66.9  \\ \toprule
        Yield & 20.5\% & 22.2\% & 14.8\% \\
	Yield w/o GLC & 92.5\% & 100\% & - \\ \bottomrule
    \end{tabular}
\label{tbl:yield_breakdown}
\end{table}

We next investigated the effect of the carbon number of each protein to the carbon yield (Fig \ref{fig:Yield_POI}B).
The same inverse qualitative trend is observed as for productivity, however it was less significant.
The proteins with the lowest carbon number had the highest yield and the higher carbon number proteins had a lower carbon yield within each case. 
As the size of the protein to be expressed increases, the productivity and carbon yield decreases, thus proteins that are too large may not be feasible for cell-free production.
Thus, we wanted to examine the most sensitive parameters for cell-free productivity and carbon yield in order to optimize CFPS systems most effectively.
%Sequence specific flux balance analysis assumes a psuedo steady state, thus intermediate metabolites cannot be accumulated within the cell-free extract.
%In addition, ssFBA is solved by setting the production of the protein as the objective function.
%Therefore, carbon flux will travel throughout the network to optimize the flux through the protein synthesis reaction.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/Yield_POI.pdf}
\caption{CFPS carbon yield performace for eight proteins for the control (black), AA uptake w/o synthesis (off white), and AA synthesis w/o uptake (grey). A. Carbon yield (Error bars represent the 95\% CI of the ensemble). B. Mean carbon yield versus the carbon number for each corresponding protein.}
\label{fig:Yield_POI}
\end{figure}

\subsection{CFPS performance sensitivity analysis}
To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance we performed global sensitivity analysis on the productivity and carbon yield for deGFP, a representative protein (Fig \ref{fig:SI_GFP}).
In examining productivity performance (Fig \ref{fig:SI_GFP}A), the significance of transcription/translation parameters were fairly constant across all three cases, with the elongation rate by ribosomes being the most sensitive.
This showed the rate of translation is instrumental in achieving high productivity, and should be the first step to investigate in order to optimize productivity, prior to examining transcription rates, this is consistent with literature findings.
Underwood and coworkers have also shown that an increase in ribosome levels does not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
In addition, Li et al. have increased productivity of firefly lucifease by 5-fold in CFPS systems by adding and adjusting factors that affect transcription and translation such as elongation factors, ribosome recyclcing factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
In examining the substrate utilization, glucose uptake was not very important for productivity in the first two cases, but its significance increased when amino acids were removed from CFPS.
This makes sense, as amino acid synthesis can only come from glucose and became the only way to power protein synthesis.
On the other hand, when amino acid synthesis is removed, amino acid uptake became more important, as it becomes the only source of amino acids.
\begin{figure}[t!]
\includegraphics[width=1.00\textwidth]{./Figures/SI_GFP.pdf}
\caption{Total order sensitivity of deGFP productivity (A) and carbon yield (B) to specific uptake rates and transcription/translation parameters for three cases: control (black), amino acid uptake without synthesis (off white), and amino acid synthesis without uptake (gray). (Error bars represent 95\% CI of the ensemble).}
\label{fig:SI_GFP}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% The abstract environment will automatically gobble the contents
%% if an abstract is not used by the target journal.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
In this study, we used sequence specific constraints based modeling to evaluate the performance of synthetic circuits in an \emph{E.~coli} cell-free protein synthesis system.
A core \emph{E.~coli} metabolic model, consisting of glycolysis, pentose phosphate pathway, amino acid biosynthesis and degradation and energy metabolism, was then augmented with sequence specific descriptions of genetic circuits which included mechanistic models of promoter function, transcription and translation.
Thus, unlike other synthetic biology modeling efforts, sequence specific constraints based modeling explicitly couples the transcription and translation of circuit components with the availability of metabolic resources.
Model parameters for transcription and translation were taken from literature, allowing a first principles prediction of circuit performance.
We tested this approach by first simulating T7 induced CAT production and $\sigma_{70}$-induced deGFP expression; we then expanded these studies for a range of different proteins.
First principles predictions of circuit performance were consistent with measurements for a variety of cases.
Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance in terms of productivity and yield.
A sufficient energy supply with oxidative phosphorylation is instrumental for high carbon yields; in addition, the translation rate is the next step to optimize for higher productivity and yield.
Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.
\end{abstract}

@@

In this study, we used sequence specific constraints based modeling to evaluate the performance of synthetic circuits in an \emph{E.~coli} cell-free protein synthesis (CFPS) system.
A core \emph{E.~coli} cell free metabolic model, consisting of 132 metabolites and 281 reactions, was developed from literature \cite{Feist:2007aa}.
This model, which described glycolysis, pentose phosphate pathway, amino acid biosynthesis and degradation and energy metabolism, was then augmented with
sequence specific descriptions of genetic circuits which included mechanistic models of promoter function, transcription and translation.
Thus, sequence specific constraints based modeling explicitly coupled the transcription and translation of circuit components with the availability of metabolic resources.
We tested this approach by first simulating the production of CAT and deGFP under two different promoters, and then expanded this for $\sigma_{70}$-induced deGFP expression at different plasmid concentrations.
First principles predictions of circuit performance were consistent with measurements.
We then investigated the productivity and carbon yield for a range of 8 different proteins for a variety of cases.
From this, higher carbon number proteins typically had lower productivity rates and carbon yields than that of the lower carbon number proteins.
Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance, showing oxidative phosphorylation is instrumental for maintaining a high carbon yield and the translation rate for productivity.
Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell free synthetic circuits.

@@

\subsection{Model Derivation}
The goal of this work was first to construct a modeling framework to describe cell-free protein synthesis systems and to examine its performance in productivity and yield for a protein of interest.
One mathematical framework that has found wide use in modeling metabolism is constraint based models such as flux balance analysis (FBA).
FBA can predict how cells utilize nutrients to produce products by using the biochemical stoichiometry and thermodynamical feasibility under pseudo steady-state conditions.
Traditionally, FBA is used to model \textit{in vivo} processes; however, cell-free systems do not have growth associated reactions or transport through the cell membrane.
Thus, to model cell-free metabolism we constructed a cell-free stoichiometric network by removing growth associated reactions from the MG1655 reconstruction \cite{Feist:2007aa}, and incorporating amino acid synthesis and transcription/translation associated reactions \cite{Allen:2003aa} for a protein of interest to be expressed.
The network consisted of 281 reactions and 132 species.
We developed a cell-free sequence specific flux balance analysis (ssFBA) with a detailed promoter model \cite{Moon:2012aa} to examine the performance of CFPS.
We first validated the ssFBA approach by comparing simulated and measured concentrations of two proteins from two different cell-free \textit{E. coli} extract systems.
The first protein, chloramphenicol acetyltransferase (CAT), was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress} for a duration of 1 hour under glucose consumption (Fig \ref{fig:CAT_GFP_prod}A).
The second protein, dual emission green fluorescent protein (deGFP), was produced under a P70 promoter in TXTL 2.0 \textit{E. coli} extract for a duration of 8 hours under maltose consumption (Fig \ref{fig:CAT_GFP_prod}B).
The ssFBA simulations predicted the production of both proteins for the duration of both CFPS batch reactions.
Uncertainty in experimental factors such as RNA polymerase, ribosome concentrations, elongation rates and the upper bounds for oxygen and glucose consumption rates did not alter the qualitative performance of the model.
Thus, the metabolic network and molecular description of trancription and translation were consistent with experimental measurements.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/CAT_GFP_prod.pdf}
\caption{Sequence specific flux balance analysis of protein production vesus time. A. CAT production under a T7 promoter in CFPS \textit{E. coli} extract for 1 hr under glucose consumption. B. deGFP production under a P70 promoter in TXTL 2.0 \textit{E. coli} extract for 8 hr under glucose consumption. 95\% CI (blue region) over the ensemble of 100 sets.}
\label{fig:CAT_GFP_prod}
\end{figure}

@@

\subsection{CFPS Productivity}
Our next goal was to examine the performance of CFPS productivity for eight different proteins under three different cases.
Each of the proteins was produced under a P70 promoter, except for CAT which was produced under a T7 promoter.
In all cases, CFPS was supplied with glucose.
In the first case, CFPS was supplied with amino acids, and the system was allowed to synthesize amino acids (control).
In the second case, CFPS was supplied with amino acids, but the amino acid synthesis was turned off (AA uptake w/o synthesis).
We shut off these synthesis reactions since during the cell-free extract preparation the cells are often supplied with amino acids; thus, the enzymes responsible for amino acid synthesis would not be present.
In the third case, CFPS was not supplied with amino acids, but the system could synthesize them (AA synthesis w/o uptake).
Eight different proteins, ranging in size, were selected to evaluate CFPS performance: bone morphogenetic protein 10 (BMP10), chloramphenicol acetyltransferase (CAT), caspase 9 (CASP9), dual emission green fluorescent protein (deGFP), prothrombin (FII), coagulation factor X (FX), fibroblast growth factor 21 (FGF21), and single chain variable fragment R4 (scFvR4).
We used ssFBA to estimate the productivity for each of these proteins for each case (Fig \ref{fig:Prod_POI}A).
The first two cases had very similar performance, with the second case (without amino acid synthesis) having a slightly higher productivity for most proteins. 
%The second case (without amino acid synthesis) showed the highest productivity for each of the proteins, however the control case had very similar performance.
This shows the system had sufficient substrates to power the system and synthesize each protein.
The third case had the lowest productivity for each protein; thus, the addition of amino acids to CFPS extract is important for maintaining a relatively high productivity.
The drop in productivity in the third case is due to the fact that glucose is the only substrate supplied and must be used to provide the necessary energy requirements for transcription and translation as well as synthesize each amino acids required for the production of the protein of interest.  
The qualitative trend of productivity for the three different cases was the same within each case; however, some proteins had higher productivity than others.
For instance, in the second case, FGF21 (fibroblast growth factor 21) had a productivity of 17 ($\mu$M/h) whereas FII (prothrombin) had a productivity of 3 ($\mu$M/h).
To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig \ref{fig:Prod_POI}B).
The proteins with the highest productivity had the lowest carbon number, whereas proteins with low productivity had higher carbon numbers.
This inverse trend was due to the fact that larger proteins require more amino acids and substrates to assemble them, resulting in lower productivity.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/Prod_POI.pdf}
\caption{CFPS productivity performace for eight proteins for the control (black), AA uptake w/o synthesis (off white), and AA synthesis w/o uptake (grey). A. Productivity normalized to the specific glucose uptake rate (Error bars represent the 95\% CI of the ensemble). B. Mean productivity versus the carbon number for each corresponding protein.}
\label{fig:Prod_POI}
\end{figure}

\subsection{CFPS Carbon Yield}
Following the same outline as in examining the productivity, we calculated the carbon yield for each protein (Fig \ref{fig:Yield_POI}A).
The same trends followed, where the case without amino acid synthesis showed the highest carbon yield and the control case had comparable performance.
The third case (with no amino acid uptake) had the lowest yields; this is most likely because glucose is utilized to synthesize the necessary amino acids for each protein as well as power the system. 
To determine the carbon contribution from each substrate (glucose and amino acids) we examined the carbon flux going toward the production of deGFP for all three cases (Table \ref{tbl:yield_breakdown}).
For the first case, the system relied on a mixture of glucose and some amino acids for the production of deGFP with a carbon yield of 20.5\%.
Once amino acid synthesis was removed from the network (second case), each amino acid was utilized and the carbon yield increased to 22.2\%.
Only the necessary amount of amino acids were used for the production of deGFP; thus, it may be hypothesized that all the glucose was used to power CFPS and does not contribute to the carbon yield.
If that is the case, the carbon yield without glucose contribution is 100\%, whereas for the first case it is 92.5\%.
Finally for the third case (without amino acids supplemented), the carbon yield was reduced to 14.8\% and the system used about twice the amount of glucose as in the first two cases.
In this case, glucose is being used to synthesize amino acids and provide the necessary energy requirements to power transcription and translation; this trend was seen across all proteins except for CAT.
Interestingly, CAT carbon yield showed similar performance for all three cases; however, it still had a lower yield than the first two cases.
Each protein had different energy requirements for transcription and translation depending on its sequence.
In the case of CAT, energy requirements for transcription or translation were low enough that its carbon yield was not hindered in the third case as significantly as for the other proteins.

\begin{table}[t!]
\centering
    \caption{Carbon contribution from glucose and each amino acid toward deGFP production for three cases: control, amino acid uptake without synthesis, and amino acid synthesis without uptake.}
    \renewcommand{\arraystretch}{0.9}
    \begin{tabular}{lccc} \toprule
        \parbox{3.3 cm}{Carbon \phantom{abcdefghi} Produced (mM/hr)} & Control & \parbox{2.6 cm}{AA uptake \phantom{ai} w/o synthesis} & \parbox{2.5 cm}{AA synthesis w/o uptake} \\ \midrule
        deGFP & 9.8 & 9.6 & 9.9  \\ \toprule
        \parbox{3.3 cm}{Carbon \phantom{abcdefghi} Consumed (mM/hr)} & \\ \midrule
        GLC & 37.3 & 33.7 & 66.9 \\
        ALA & 0.0 & 0.2 & - \\
        ARG & 0.3 & 0.3 & - \\
        ASN & 0.5 & 0.4 & - \\
        ASP & 0.4 & 0.6 & - \\
        CYS & 0.2 & 0.1 & - \\
        GLU & 2.1 & 0.6 & - \\
        GLN & 0.4 & 0.3 & - \\
        GLY & 0.3 & 0.3 & - \\
        HIS & 0.5 & 0.5 & - \\
        ILE & 0.0 & 0.6 & - \\
        LEU & 1.0 & 1.0 & - \\
        LYS & 1.0 & 0.9 & - \\
		MET & 0.0 & 0.2 & - \\
        PHE & 1.0 & 0.9 & - \\
        PRO & 0.4 & 0.4 & - \\
        SER & 0.0 & 0.2 & - \\
        THR & 1.0 & 0.5 & - \\
        TRP & 0.1 & 0.1 & - \\
        TYR & 0.8 & 0.8 & - \\
        VAL & 0.6 & 0.7 & - \\ \hline
        Sum & 47.9 & 43.3 & 66.9 \\ \toprule
        Yield & 20.5\% & 22.2\% & 14.8\% \\
	Yield w/o GLC & 92.5\% & 100\% & - \\ \bottomrule
    \end{tabular}
\label{tbl:yield_breakdown}
\end{table}

We next investigated the effect of the carbon number of each protein on the carbon yield (Fig \ref{fig:Yield_POI}B).
The same inverse qualitative trend is observed as for productivity, but it was less significant.
The proteins with the lowest carbon number had the highest yield and the higher carbon number proteins had a lower carbon yield within each case. 
As the protein length increased, the productivity and carbon yield decreased, suggesting that large proteins may not be feasible for cell-free production.
We thus wanted to examine the most sensitive parameters for cell-free productivity and carbon yield in order to optimize CFPS performance.
%Sequence specific flux balance analysis assumes a psuedo steady state, thus intermediate metabolites cannot be accumulated within the cell-free extract.
%In addition, ssFBA is solved by setting the production of the protein as the objective function.
%Therefore, carbon flux will travel throughout the network to optimize the flux through the protein synthesis reaction.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./Figures/Yield_POI.pdf}
\caption{CFPS carbon yield performace for eight proteins for the control (black), AA uptake w/o synthesis (off white), and AA synthesis w/o uptake (grey). A. Carbon yield (Error bars represent the 95\% CI of the ensemble). B. Mean carbon yield versus the carbon number for each corresponding protein.}
\label{fig:Yield_POI}
\end{figure}

\subsection{CFPS performance sensitivity analysis}
To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance we performed global sensitivity analysis on the productivity and carbon yield for deGFP, a representative protein (Fig \ref{fig:SI_GFP}).
In examining productivity performance (Fig \ref{fig:SI_GFP}A), the significance of transcription/translation parameters was fairly constant across all three cases, with the elongation rate by ribosomes being the most significant.
This showed that translation rate is instrumental for productivity, and should be the first step investigated in its optimization, prior to examining transcription rates.
This is consistent with literature findings.
Underwood and coworkers have also shown that an increase in ribosome levels does not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
In addition, Li et al. have increased productivity of firefly luciferase by 5-fold in CFPS systems by adding and adjusting factors that affect transcription and translation such as elongation factors, ribosome recyclcing factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
In examining the substrate utilization, glucose uptake was not very important for productivity in the first two cases, but its significance increased when amino acids were removed from CFPS.
This makes sense, as amino acid synthesis from glucose became the only way to power protein synthesis.
On the other hand, when amino acid synthesis was removed, amino acid uptake became more important, as it was the only source of amino acids.
\begin{figure}[t!]
\includegraphics[width=1.00\textwidth]{./Figures/SI_GFP.pdf}
\caption{Total order sensitivity of deGFP productivity (A) and carbon yield (B) to specific uptake rates and transcription/translation parameters for three cases: control (black), amino acid uptake without synthesis (off white), and amino acid synthesis without uptake (gray). (Error bars represent 95\% CI of the ensemble).}
\label{fig:SI_GFP}
\end{figure}

When considering carbon yield performance (Fig \ref{fig:SI_GFP}B), the substrate and oxygen uptake fluxes became much more important while the sensitivity to transcription/translation parameters decreased slightly.
The transcription/translation parameters had the same trend as for productivity, where the ribosome elongation rate was the most sensitive compared to the other transcription/translation parameters and showed significance across all cases.
Thus, in investigating CFPS performance in terms of productivity and yield, the ribosome elongation rate is an important parameter to optimize, as has already been shown in literature \cite{2005_underwood_biotech, 2014_li_PlosOne}.
Glucose and oxygen uptake showed significant importance for carbon yield since they determined the mechanism of energy generation, which is critical for efficient protein production.
Meanwhile, productivity was determined primarily by the rate of the most downstream steps, transcription and translation.
Glucose is required for both amino acid synthesis and efficient energy generation, both of which were important for a high yield.
This resulted in a tradeoff between energy generation to power transcription/translation and amino acid synthesis required to assemble the protein of interest.
Across all three cases, substrate utilization (amino acid uptake and glucose) showed to be the next important as these substrates contributed to the carbon yield of deGFP.
Interestingly, in the first two cases, oxygen uptake also had a significant effect on carbon yield.
This is most likely due to the importance of oxidative phosphorylation for efficient energy generation.
With high oxidative phosphorylation, carbon substrates were not wasted to meet the necessary energy requirements of CFPS; instead, carbon was directed toward the production of deGFP.
However, in the third case (with no amino acid uptake), oxygen uptake significance decreased, while glucose uptake became the most siginificant.
When amino acids were not supplied in CFPS, glucose was the only carbon source for amino acid synthesis and energy generation. 
The same tradeoff was observed between utilizing glucose for amino acid synthesis and efficient energy generation; the significance of oxygen was irrelevant compared to that of glucose uptake. 
In addition, NADH was required to power oxidative phosphorylation; however, the demand for NADPH for amino acid syntheis was higher, since amino acids were required for the production of the protein of interest. 
NADH was thus drawn away from oxidative phosphorylation to be interconverted to NADPH.
This decreased the significance of oxygen uptake, since there was not enough NADH to power oxidative phosphorylation.
Jewett and coworkers have reported oxidative phosphorylation still operating in cell-free systems, and that yield decreased from 1.5-fold to 4-fold when oxidative phosphorylation reactions were knocked out\cite{Jewett:2008aa}.
However, it is unknown how active oxidative phosphorylation is compared to that of \textit{in vivo} systems.
Thus, to investigate this further we compared deGFP carbon yield to oxidative phosphorylation flux (Fig \ref{fig:oxphos_yield}).
Interestingly, oxidative phosphorylation has a somewhat linear correlation with carbon yield for the first two cases (R\textsuperscript{2}=0.77 for the control, R\textsuperscript{2}=0.79 for the second case), whereas for the third case, this effect was not as prevalent (R\textsuperscript{2}=0.46).
\begin{figure}[t!]
\includegraphics[width=0.7\textwidth]{./Figures/OxPhos_yield.pdf}
\caption{An ensemble of 1000 ssFBA solutions for deGFP carbon yield versus oxidative phosphorylation flux for three cases: control (black), amino acid uptake without synthesis (off white), and amino acid synthesis without uptake (gray).}
\label{fig:oxphos_yield}
\end{figure}

When the system can utilize glucose for energy generation, the increase in oxidative phosphorylation has a positive effect on carbon yield.
However, when glucose is used to synthesize amino acids and power CFPS, there is a limit on how much oxidative phosphorylation can effect carbon yield due to the limitation of NADH to power oxidative phosphorylation.
Thus, it is interesting to see how much of the energetic needs of the system are met by oxidative phosphorylation and where the remaining energy is coming from.
To investigate the differences in productivity and carbon yields, we compared the flux distributions for deGFP predicted by ssFBA simulations for the three different cases (Fig.\ref{fig:flux}).
\begin{figure}[t!]
\includegraphics[width=0.78\textwidth]{./Figures/flux.pdf}
\caption{deGFP production flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, NADPH/NADH transfer, and oxidative phosphorylation. Flux across ensemble (mean $\pm$ standard deviation) normalized to glucose uptake flux. Flux distribution for three different cases: control (top row), amino acid uptake without synthesis (second row), and amino acid synthesis without uptake (third row). Bold metabolites are allowed to accumulate; grey arrows lead to amino acid synthesis.}
\label{fig:flux}
\end{figure}
